\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{fancyvrb}
\usepackage{minted}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumitem}  %this permits the enumeration items to flow past sections


\title{Algorithms - Assignment 7}
\date{Due }

\begin{document}
\maketitle

\begin{enumerate}
    \item[(1)] Let $A$ be an array of integers. We wish to find a subarray $A[i:j]$ (i.e. a subarray of consecutive elements from $A$, sometimes referred to as a 
    emph{contiguous subsequence}) such that when we sum those entries, we get the largest possible sum. For example, if $A=[-2, 1, -3, 4, -1, 2, 1, -5, 4]$, then a contiguous subarray with the maximum sum is $[4,-1,2,1]$, whose elements sum to $6$.
    \begin{itemize}
        \item[(a)] Describe and implement a brute-force solution to the problem. What is the runtime? 
        \item[(b)] Describe a dynamic programming solution which operates in linear time. Your description needs to clearly define the subproblems, express a recurrence relation between the problem and those subproblems, and clearly state what is being done in order to construct the solution (i.e. what we are maximizing, minimizing, etc). You should also justify the runtime. 
        \item[(c)] Implement the solution in Python. Your solution can be top down or bottom up, and doesn't need to return the actual subsequence, just the maximum sum. 
    \end{itemize}
    \item[(2)] 
\end{enumerate}

\end{enumerate}


\end{document}